---
title: "Formal Demography Workshop: In-Class Exercises on `Replacement Migration' (Part 2)"
author: "Joshua R. Goldstein and Dennis Feehan"
output: html_notebook
---

# Overview of Part 2

Now we're ready to actually try to replicate the UN's _Replacement Migration_
method for projecting populations with net migration.

Our goals in this part are: 

1. To implement population projections with time-varying Leslie matrices

2. To operationalize the UN's approach to "replacement migration" with one of various _objectives_


# Time-varying Leslie matrix projection

We'll implement the UN assumptions abut fertility and mortality from
1995 to 2050, the time span of the Replacement Migration report. This
means that our population projection without migration will have the
form:

$$
{\bf K}_{t+5} = {\bf A}_t {\bf K}_t,
$$
where ${\bf K}$ is a column vector of 5 year age-group counts and
${\bf A}_t$ is a Leslie matrix built with the period fertility and
mortality rates specified for time $t$.

Be patient, please. We'll look at how to include migration in a bit :)

## Implementation

Our approach will be to use some pre-built functions that I (Josh) wrote up.

```{r}
leslie <- function(x= seq(0, 100, 5), nFx, nLx)
{
    ## defaults to 5 year age groups
    if(length(nFx) != length(x))
        cat("nFx wrong length\n")
    if(length(nLx) != length(x))
        cat("nLx wrong length\n")

    ## define next age group
    nFxpn = c(nFx[-1], 0)
    nLxpn = c(nLx[-1], 0)

    ## build leslie matrix (0.4886 is proportion female at birth)
    firstrow = (nLx[1]/2) * (nFx + nFxpn * nLxpn/nLx) * 0.4886
    subdi = nLxpn/nLx

    ## put subdi and firstrow together
    subdi.mat <- diag(subdi[-length(subdi)])
    A <- rbind(firstrow, cbind(subdi.mat,0))

    ## print out intrinsic rate of natural increase "r"
    ## as a check
    lambda = Re(eigen(A)$val[1])
    r = log(lambda)/5
    cat("r = ")
    cat(round(r,4))
    cat("\n")
    return(A)
}

leslie_for_period <- function(period, nFxt, nLxt)
{
    ## this indexes the matrices with the particular period
    ## and fixes the age groups for nFx and nLX
    ## usage: 
    ##.   A_2015 <- leslie_for_period(2015, nFxt = nFxt, nLxt = nLxt)
    x = seq(0, 100, 5)
    ##
    nFx_raw = nFxt[,paste(period)]
    nFx = c(rep(0, 2), nFx_raw, rep(0, 10)) ## fill out to all ages
    names(nFx) <- x
    ##
    nLx_raw = nLxt[,paste(period)]
    ## x_mort = c(0, 1, seq(5, 100, 5))
    nLx = c(sum(nLx_raw[1:2]), nLx_raw[-(1:2)]) ## collapse childhood to 0-4
    names(nLx) <- x

    A <- leslie(x = x, nFx, nLx)
    return(A)
}
```

## Testing on Italy

Re-read in the data and format it
```{r}
## nFxt
my_location = "Italy"
dt <- fread("./nFx.csv")
mydt <- dt[Location == my_location]
nFxt <- mydt[, xtabs(nFx ~ x + t)]
## nLxt
dt <- fread("./nLx.csv")
mydt <- dt[Location == my_location]
nLxt <- mydt[, xtabs(nLx ~ x + t)]
## nKxt
dt <- fread("./nKx.csv")
mydt <- dt[Location == my_location]
nKxt <- mydt[, xtabs(nKx ~ x + t)]
```

Project Italy without migration from 1995 to 2050
```{r}
## initialization
x = seq(0, 100, 5)
t = seq(1995, 2050, 5)
nKxt_hat <- matrix(NA, length(x), length(t)) ## "hat" means projected
dimnames(nKxt_hat) <- list(x,t)
nKxt_hat[,"1995"] <- nKxt[,"1995"] ## starting pop is observed 1995

for (i in 1:(length(t)-1)) ## last period starts in 2045
{
	cat(paste("t = ", t[i], " ")) ## for nicer output with dates ...
	## get period specific leslie matrix
    At = leslie_for_period(period = t[i], nFxt=nFxt, nLxt = nLxt)
    ##
    ## project 1 time step ahead
    nKxt_hat[,i+1] <- At %*% nKxt_hat[,i]
}
```

Notice the intrinsic growth rates of the matrices. Are they varying?
Are they positive or negative? Are they growing or shrinking? Is this
consistent with the fertility forecast we saw in Part 1?


Let's look at total pop size and compare it to the medium forecast
```{r}
Kt_hat = colSums(nKxt_hat) ## what we projected w/o migration
Kt_medium = colSums(nKxt) ##  what UN projected w/ a bit of migration

plot(names(Kt_medium), Kt_medium/million, ylim = range(Kt_medium)*c(.9, 1.1)/million, ylab = "millions",
     xlab = "period", main = paste(my_location, "projected population sizes"))
lines(names(Kt_hat), Kt_hat/million, lwd = 3, col = 'red')
legend("topright", legend = c("UN Medium (w/migration)", "Our projection (w/o migration)"),
       col = c("black", "red"), pch = c(1, -1), lty = c(-1, 1))
```

We see that medium projection is quite a bit higher --- presumbably
because it already includes some net immigration.

# Replacement migration

Now we're ready to include migration. The UN's approach is to add in
migrants at every time step whenever they are needed in order to meet
the specified "objective". The objectives range from keeping total
population size constant (scenario 2) to keeping the
old-age-support-ratio constant (scenario 5). (You should check to see
if I got the scenario numbers correct, and also to see what the other
scenarios are.)

The mathematical form of the projection is now

$$
{\bf K}_{t+5} = {\bf A}_t {\bf K}_t + {\bf G}_t,
$$

where ${\bf G}_t$ is a column vector of the counts of immigrants at
each age needed to meet the "objective" at time $t+5$.

A further restriction is that the age structure of net immigration is
assumed to be known and fixed as the average of net migration
schedules in the US, Canada, and Australia. Here is the schedule
copied from the report

```{r}
ngx = c(8.63, 9.26, 8.30, 8.31, 12.25, 16.65, 12.06,
        7.68, 4.51, 2.76, 2.40, 2.25, 2.05, 1.55, 0.77, 0.56, 0.02, ## thru 85
        0, 0, 0, 0)/100 ## padded zeros
names(ngx) <- x
sum(ngx)
```


```{r}
par(mfrow=c(1,2))
barplot(ngx, horiz=TRUE, col = "blue", 
        xlab = "Proportion",
        main = "Net migration age schedule")
ncx <- nKxt[,"1995"]/sum(nKxt[,"1995"])
barplot(ncx, horiz = TRUE, col="red",
        main = "Age distribution\nof initial population",
        xlab = 'Proportion')
```



You can see that $ngx$ sums to 1.0, and so we can multiply it by the number of net immigrants to get the vector ${\bf G}$.

The logic of each projection step is 

1. Project 1-step without migration

2. Calculate how many net migrants during that 1-step are needed to
   meet objective

3. Add that number to the projected population without migration

4. Iterate the above steps, keeping track of total population by age
   and by total immigrants by age.

There's one additional twist. The UN objectives involve keeping some
quantity constant, but in some cases the quantity "improves" for some
time before getting "worse". In this case, the UN doesn't start
migration until things threaten to get worse -- that is, they try to
keep things constant after they reach the "best" state. This means
that there's an additional step of checking whether things are
"getting worse" before checking on how many migrants are needed. (This
will turn out to produce some hard-to-interpret results in some
special cases.)

## Implementation in R

Our objective will be to keep the OASR constant, so it's useful to have a function that turns a vector of age counts into this ratio
```{r}
get_old_age_support_ratio <- function(nKx, x = seq(0, 100, 5))
{
    working = sum(nKx[x %in% 15:64])
    old = sum(nKx[x >= 65])
    young = sum(nKx[x < 15])
    oasr = working / old
    return(oasr)
}
```

Our function that calculates how many migrants are needed is

```{r}

## Note: bad news is the code here is rather difficult to
##   understand. good news is that it is fairly generic and you should
##   be able to use without understanding.

get_migrants <- function(summary_function, 
                         nKx_target, 
                         nKx_nomig,
                         migration_schedule)
{
    ## usage: get_migrants(summary_function = get_old_age_support_ratio,
    ##                    nKx_target = nKxt_hat[,i],
    ##                    nKx_nomig = nKx_nomig,
    ##                    migration_schedule = ngx)
    # this function returns the counts of migrants that would be
    ## needed that would keep some objective constant

    ## if this_value is not "worse" than previous value, return 0 migrants
    target_value <- summary_function(nKx_target)
    this_value <- summary_function(nKx_nomig)
    if(this_value >= target_value)
    {
        ## no migration
        mig_x = rep(0, length(migration_schedule))
    }
    if(this_value < target_value)
    {
        ## yes, migration 
        f <- function(theta, nKx_target, nKx_nomig)
        {
          ## a function that calculates how different summary values are 
          ## with some number of migrants and the ideal of the target
            summary_function(nKx_target) - summary_function(nKx_nomig + theta * migration_schedule)
        }
        million = 10^6
        billion = 1000*million
        ## now we find the count of immigrants "theta" that maintains target.
        ## f() returns 0 when we have the right number.
        ## uniroot() search across theta from 1 to 10 billion and finds theta that makes f() == 0
        out = uniroot(f = f, nKx_target = nKx_target, nKx_nomig = nKx_nomig,
                      interval = c(1, 10*billion))
        number_of_migrants = out$root
        mig_x = number_of_migrants * migration_schedule  ## vector of counts by age
    }
    return(mig_x)
}
```

Now let's do the projection replacement migration

```{r}
## initialization
x = seq(0, 100, 5)
t = seq(1995, 2050, 5)
nKxt_hat <- matrix(0, length(x), length(t)) ## "hat" means projected
dimnames(nKxt_hat) <- list(x,t)
nKxt_mig_only <- nGxt <- nKxt_hat ## initialize as 0s
nKxt_hat[,"1995"] <- nKxt[,"1995"] ## initialize starting pop with observed in 1995
for (i in 1:(length(t)-1)) ## last projection period starts in 2045, not 2050
{
    ## get period specific leslie matrix
    At = leslie_for_period(period = t[i], nFxt=nFxt, nLxt = nLxt)
    ##
    ## project 1 time step ahead without migrants
    nKx_nomig <- At %*% nKxt_hat[,i]
    ##
    mig_x = get_migrants(summary_function = get_old_age_support_ratio,
                         nKx_target = nKxt_hat[,i],
                         nKx_nomig = nKx_nomig,
                         migration_schedule = ngx)
    nKxt_hat[,i+1] = nKx_nomig + mig_x
    nGxt[,i] <- mig_x
    ## also keep track of migrant pop and descendants for graph
    nKxt_mig_only[,i+1] = At %*% nKxt_mig_only[,i] + mig_x ## + nGxt[,i]
}

## let's check to make sure we're meeting the objective by seeing if
## in this case the old age support ratio stays constant
objective_test = apply(nKxt_hat, 2, get_old_age_support_ratio)
print(cbind(objective_test))
```

Looks good! We've kept (in the case of Italy) the old-age-support
ratio at a value of 4.09

Now let's visualize our results by looking at the age-pyramids,
separated by the descendants of "natives" (as of 1995) and the
descendants of immigrants (post 1995).

```{r, }
nKxt_native_only = nKxt_hat - nKxt_mig_only
my_t.vec <- t
par(mfrow = c(4,3), mar = c(2,2,1,1))
## par(mfrow = c(1,1))
for (i in 1:length(my_t.vec))
{
    my_t = paste(my_t.vec[i])
    barplot(nKxt_hat[,my_t]/million, horiz = T, col = "white", xlab = "millions")
    barplot(nKxt_native_only[,my_t]/million, col = "black", horiz = T, add = T)
    Kt = sum(nKxt_hat[,my_t])
    title(paste(my_location, my_t, "\n", round(Kt/million), "million"),
          cex.main = .7)
}
```

The age-pyramid in "black" is the descendants of the population
present in Italy in 1995 and the population in "white" is the
descendants of the population of migrants (including the migrants
themselves) arriving after 1995.

Also, note, the x-axis is changing and that the size of the population
is increasing a lot, quadrupling from 1995-2050.

## Immigration totals

We know produce a table like the UN did, with migrants each period and
annual migration, along with the total

```{r}
Gt_per_5_years = colSums(nGxt)/million
Gt_per_year = Gt_per_5_years/5
G_total = sum(Gt_per_5_years)
Kt = colSums(nKxt_hat)/million
net_migration_rate = Gt_per_year/Kt
print('note: these totals are millions of migrants')
cbind(Gt_per_5_years = round(Gt_per_5_years,1),
      Gt_per_year = round(Gt_per_year, 1),
      net_migration_rate = round(net_migration_rate, 3))
print("total millions from 1995-1950")
print(G_total)
```

We can see that the net migration rate is something on the order of
2% of the total population. Big but not implausible.

We could compare this to an approximation of the crude birth rate that
we obtain by dividing the youngest age-group by 5 to get the
approximate number of births. (Note: the youngest age-group also
contains some immigrants, but these are also included in the
denominator of the population, so might not be that much of an issue.)

```{r}
Bt_approx = (nKxt_hat["0",]/5)/million
crude_birth_rate_approx = Bt_approx/Kt
cbind(net_migration_rate = round(net_migration_rate, 3),
   	 crude_birth_rate_approx = round(crude_birth_rate_approx, 3))
```

So, in Italy, "replacement migration" to keep the old-age-support
ratio constant at 4.1 workers per elderly person would imply annual
net immigration at about 2 to 4 times the rate of births.

# Exercises

1. Repeat the replacement migration estimates that we just did for Italy, this time using "Republic of Korea."

* What is the level of old-age support ratio that is being maintained?

* How many total migrants are needed? 

* How much would the South Korean population grow?

* What would be the annual crude migration rate?

* What would you say is the main reason that so many migrants are needed?

2. What would be some other "objectives" to implement? Look up the UN
replacement migration scenarios -- and make up a new one yourself?

(You don't have to do any coding for now. We'll try that later)

3. What do you think would happen in South Korea if we kept the
old-age support ratio constant at its 2020 levels (instead of 1995)?
Would this imply higher or lower rates of "replacement migration"?

(You might want to plot the age-pyramids of the UN projection
"nKxt" for 1995 and 2025 to inform you answer).
(But you can also do this without coding  -- just give your reasoned guess.)


# Congratulations!

> You've finished the 2nd part of our "Replacement Migration" exercise.
> The last part is to change some of the assumptions and see how much the results change.