---
title: "Formal Demography Workshop: In-Class Exercises on `Replacement Migration' (Part 3)"
author: "Joshua R. Goldstein and Dennis Feehan"
output: html_notebook
---


# Overview of Part 3

Here we experiment with some alterative objectives and assumptions to
see how this influences "replacement migration". We also provide
details for your team project.


Our goals are: 

1. To see if there was something special about 1995 that affected our
estimates. We'll do this by trying to maintain the old-age support
ratio observed in 2020 rather than 1995.

2. To change the objective to keeping the size of the working
population constant. This is on of the UN's scenarios. This will also
allow us to see how to code our own, customized objective.

3. To provide instructions for the team mini-project.

# Changing the starting state that is the target of "replacement"

Recall from Part 1 that the old age support ratio from Italy in 1995
was largely influenced by fertility that occurred many decades
earlier.

If we shift our sights a bit, and instead try to maintain the old-age
support ratio seen today (in 2020), then the level might be lower and
it might be easier to maintain.

We'll implement this change by starting our projection with the 2020
population age structure, but for comparability -- and easy of coding
-- we'll use the same fertility and mortality rates from 1995 to 2050
to do our projection. This will make all of the differences we find
attributable to the starting age structure -- and not to differences
in fertility or mortality.

We can re-cycle our earlier code with one change, which I comment
below.

```{r}
## initialization
x = seq(0, 100, 5)
t = seq(1995, 2050, 5)
nKxt_hat <- matrix(0, length(x), length(t)) ## "hat" means projected
dimnames(nKxt_hat) <- list(x,t)
nKxt_mig_only <- nGxt <- nKxt_hat ## initialize as 0s
nKxt_hat[,"1995"] <- nKxt[,"2020"] ## <<<--- we changed initial population to have 2020 age-structure
## projection
for (i in 1:(length(t)-1)) ## last projection period starts in 2045, not 2050
{
    ## get period specific leslie matrix
    At = leslie_for_period(period = t[i], nFxt=nFxt, nLxt = nLxt)
    ##
    ## project 1 time step ahead without migrants
    nKx_nomig <- At %*% nKxt_hat[,i]
    ##
	 	   mig_x = get_migrants(summary_function = get_old_age_support_ratio,
                         nKx_target = nKxt_hat[,i],
                         nKx_nomig = nKx_nomig,
                         migration_schedule = ngx)
    nKxt_hat[,i+1] = nKx_nomig + mig_x
    nGxt[,i] <- mig_x
    ## also keep track of migrant pop and descendants for graph
    nKxt_mig_only[,i+1] = At %*% nKxt_mig_only[,i] + mig_x ## + nGxt[,i]
}
```

Let's see what level of the old-age support ratio we are maintaining

```{r}
## let's check to make sure we're meeting the objective by seeing if
## in this case the old age support ratio stays constant
objective_test = apply(nKxt_hat, 2, get_old_age_support_ratio)
print(cbind(objective_test))
```

We can see that it constant at about 2.7, after 2025, but strangely it
was constant at a slightly higher level before 2025. This probably has
something to do with the dynamics over time as the rather irregular
age-pyramid ages. During the projection there is a actually a year
when the OASR went up ("2020") and there was zero migration.

```{r, }
nKxt_native_only = nKxt_hat - nKxt_mig_only
my_t.vec <- t
par(mfrow = c(4,3), mar = c(2,2,1,1))
## par(mfrow = c(1,1))
for (i in 1:length(my_t.vec))
{
    my_t = paste(my_t.vec[i])
    barplot(nKxt_hat[,my_t]/million, horiz = T, col = "white", xlab = "millions")
    barplot(nKxt_native_only[,my_t]/million, col = "black", horiz = T, add = T)
    Kt = sum(nKxt_hat[,my_t])
    title(paste(my_location, my_t, "\n", round(Kt/million), "million"),
          cex.main = .7)
}
```

Compare to the projections we did early trying to The age-pyramid in
"black" is the descendants of the population present in Italy in 1995
and the population in "white" is the descendants of the population of
migrants (including the migrants themselves) arriving after 1995.

Rather than quadrupling the population does not quite double by 2050. 

## Immigration totals


```{r}
Gt_per_5_years = colSums(nGxt)/million
Gt_per_year = Gt_per_5_years/5
G_total = sum(Gt_per_5_years)
Kt = colSums(nKxt_hat)/million
net_migration_rate = Gt_per_year/Kt
print('note: these totals are millions of migrants')
cbind(Gt_per_5_years = round(Gt_per_5_years,1),
      Gt_per_year = round(Gt_per_year, 1),
      net_migration_rate = round(net_migration_rate, 3))
print("total millions from 1995-1950")
print(G_total)
```

We see that the replacement migration algorithm produces two separate
waves of migration, but that the total number of migrants (about 62.5
million is about half the size of what we saw earlier, starting with
the 1995 age-structure.)



# 2. Changing the objective

Here we'll implement another of the UN's scenarios, this time bringing
in enough migrants to maintain the "working" population aged 15 to 64.

We begin by writing a function that converts a vector of age-group
counts to the total number of workers

```{r}
get_workers <- function(nKx, x = seq(0,100, 5) )
{
  workers = sum(nKx[x %in% 15:64])
  return(workers)
}
```

Now we can use the same code as before, just changing the argument
summary_function in get_migrants().

```{r}
## initialization
x = seq(0, 100, 5)
t = seq(1995, 2050, 5)
nKxt_hat <- matrix(0, length(x), length(t)) ## "hat" means projected
dimnames(nKxt_hat) <- list(x,t)
nKxt_mig_only <- nGxt <- nKxt_hat ## initialize as 0s
nKxt_hat[,"1995"] <- nKxt[,"1995"] ## initialize starting pop with observed in 1995
for (i in 1:(length(t)-1)) ## last projection period starts in 2045, not 2050
{
    ## get period specific leslie matrix
    At = leslie_for_period(period = t[i], nFxt=nFxt, nLxt = nLxt)
    ##
    ## project 1 time step ahead without migrants
    nKx_nomig <- At %*% nKxt_hat[,i]
    ##
	## 	   mig_x = get_migrants(summary_function = get_old_age_support_ratio,
    mig_x = get_migrants(summary_function = get_workers, ## <<<--- we changed summary_function
                         nKx_target = nKxt_hat[,i],
                         nKx_nomig = nKx_nomig,
                         migration_schedule = ngx)
    nKxt_hat[,i+1] = nKx_nomig + mig_x
    nGxt[,i] <- mig_x
    ## also keep track of migrant pop and descendants for graph
    nKxt_mig_only[,i+1] = At %*% nKxt_mig_only[,i] + mig_x ## + nGxt[,i]
}
```

```{r}
## let's check to make sure we're meeting the objective by seeing if
## in this case the old age support ratio stays constant
objective_test = apply(nKxt_hat, 2, get_workers)/million
print(cbind(objective_test))
```
Looks good! We've kept (in the case of Italy) the number of working age people at about 39 million.

Now let's visualize our results by looking at the age-pyramids,
separated by the descendants of "natives" (as of 1995) and the
descendants of immigrants (post 1995).

```{r}
nKxt_native_only = nKxt_hat - nKxt_mig_only
my_t.vec <- t
par(mfrow = c(4,3), mar = c(2,2,1,1))
## par(mfrow = c(1,1))
for (i in 1:length(my_t.vec))
{
    my_t = paste(my_t.vec[i])
    barplot(nKxt_hat[,my_t]/million, horiz = T, col = "white", xlab = "millions")
    barplot(nKxt_native_only[,my_t]/million, col = "black", horiz = T, add = T)
    Kt = sum(nKxt_hat[,my_t])
    title(paste(my_location, my_t, "\n", round(Kt/million), "million"),
          cex.main = .7)
}
```

Note: now the population only grows slightly, from 57 to 68 million
over 55 years.

## Immigration totals

We know produce a table like the UN did, with migrants each period and
annual migration, along with the total

```{r}
Gt_per_5_years = colSums(nGxt)/million
Gt_per_year = Gt_per_5_years/5
G_total = sum(Gt_per_5_years)
Kt = colSums(nKxt_hat)/million
net_migration_rate = Gt_per_year/Kt
print('note: these totals are millions of migrants')
cbind(Gt_per_5_years = round(Gt_per_5_years,1),
      Gt_per_year = round(Gt_per_year, 1),
      net_migration_rate = round(net_migration_rate, 3))
print("total millions from 1995-1950")
print(G_total)
```
And the number of migrants is only 20 million over the entire period.

# Discussion Questions

1. Why does "replacement migration" require so many fewer immigrants
when we start with 2020 Italy than when we started with 1995 Italy?

2. Why does changing the objective function to the number of working
age adults also seem to require so many fewer immigrants to maintain?

3. Would the later target date (2020 instead of 1995) result in lower
required "replacement migration" for all populations? If you think
not, what would be a convincing counter-example?

4. Why didn't we try the total dependency ratio [workers : (old +
young)]?

5. Is the UN using the term "replacement migration" in the way you
thought of it before? Would it be better to call it "maintenance
migration" because it's keeping something constant? In a broader
sense, does bringing in a lot of migrants involve keeping the status
quo?

6. Does the critique that we (perhaps) voiced in answering Q5 lead us
to think of other alternative ways to define "replacement migration"
or to measure the quantity that bringing in migrants is supposed to
maintain?


# 3. Instructions for the team mini-project

A. We will allocate two countries to each team. Each team will then
calculate (i) the total number of migrants and (ii) the net migration
rate (in 2030) that are required to keep the old-age support ratio
constant at its 1995 level. Next, each team will recalculate again
using the 2025 starting pouplation, just as we did above for Italy. On
Thursday, we'll combine all these numbers when we meet again to get a
better understanding of how starting age-structure influences the
"replacement migration" rates. Each group should construct tables for
each that looks like this

         
Country "1 of 2"
                      Total migrants        nmr(2030)
                      1995-2050
                      (millions)
1995 pop start              A                  B
2025 pop start              C                  D

where you fill in the quantities A,B,C, and D -- and you do again for
your second country.

B. Each team should come up with its own objective function and
implement this to see how it affects the number of replacement
migrants needed.

Note 1: for ease of implementation with our existing code, it is best
if this function can be calculated based on period age-structure --
that is, if it does not depend on the age-structure or other
quantities in future or past periods.

Note 2: if the summary statistic/quantity you are calculating is such
that less is better (e.g., you want to prevent excessive aging based
on a measure like the mean age) then you will want your objective
function to return -1 times our outcome in order for the existing code
to work correctly.

Some possibilities might include

* the fraction of the entire population that is over age 65

* the number of children (under age 15) in the population

* anything else you want to try!

Please report the same summary statistics as in part A. 

I would recommend you try applying your new objective to Italy (which
we're all now familiar with) and to at least one other country that
you find interesting (easiest might be one of the countries we assigned.).

Your result can look like this:

Country "Italy"
                      Total migrants        nmr(2030)
                      1995-2050
                      (millions)
Benchmark 
(1995 pop start,
OASR objective)             A                  B
Your new objective
(1995 pop start)            C                  D

and the same for a country of your choosing. 


# Congratulations!

> You've finished all the content we have prepared for you! If you are
>  reading this before you have started your group project, then
>  congratulations on getting this far -- and good luck working with
>  your group. If you're only seeing this at the end of your group
>  project, then an even bigger congratulations: you're done!