---
title: "Formal Demography Workshop: In-Class Exercises on `Replacement Migration'"
author: "Joshua R. Goldstein and Dennis Feehan"
output: html_notebook
---


# Overview

The UN issued an influential report on _Replacement Migration_ in
2000, concluding that

> Maintaining potential support ratios at current levels through
>  replacement migration alone seems out of reach, because of the
>  extraordinarily large numbers of migrants that would be required.

For South Korea, the report contains the following astounding conclusion:

> The number of immigrants needed to maintain the potential support
>  ratio at its 1995 level (scenario V) is 110 times the size of the
>  current national population, and equal approximately to the current
>  total population of the world.

The goals of our project this year is to revisit this report. Along the way, we will:

1. Use stable population theory (without migration) to understand the
properties and implications for the future of currently observed
population pyramids.

2. Understand the UN's definition of "replacement migration" and to
produce estimates based on this definition.


3. Measure how sensitive the UN's estimates are to the starting date
of the projection.

4. Allow your team to specify your own population-related goal and to
estimate how much migration would be needed to satisfy that.

# Organizational Structure

* This document ("part 1") is about the data and starting age
  structure of the UN's _Replacement Migration_ report.  You should
  read through Part 1 before the workshop -- and you should also try
  running all of the code (and take a first pass at understanding
  it). We will review this whole document as part of the workshop on
  Wednesday.

* The next document ("part 2") implements the UN's replacmenent
  migration "algorithm". It tries to replicate the results of this report for Italy.

* Finally, the last document ("part 3") shows what happens if vary
  some of the assumptions of the UN's approach and provides a
  description of your the you will do for Thursday morning in small teams.


# Our data "ingredients"

We use the UN's schedules of fertility, mortality, and
migration. We're using files from 2022 rather than those from the
Replacement Migration report, because the newer releases contain
complete age-specific schedules of fertility and mortality for all
years.

The format of the files is as a data.frame (or equivalently a
data.table) with all of the ages and years and countries in a single
file. Our approach will be to work with one country at a time,
organizing the rates into a Lexis-surface-like age x period table.

For our in-class example, we'll look at Italy. For your teamwork,
later on, you can feel free to use other countries, including the
extreme example of South Korea.


## Vital Rates

Here we'll use the data.table package to read in the .csv files. Then
we'll create matrices of age-by-period rates.

### Fertility

```{r}
library(data.table) ## the package we use to read in and restructure the data
dt <- fread("./nFx.csv") ## we'll use the name "dt" as a placeholder
## restrict to one country
my_location = "Italy" ## you can change this to do a different country
mydt <- dt[Location == my_location] ## restricts data to this country
head(mydt) ## to see the structure of the raw data
 ## finally we create a matrix using the xtabs() functions with formula interface
nFxt <- mydt[, xtabs(nFx ~ x + t)] ## produces a table with rows age x and columns period t.
print(nFxt[1:5, 1:5])
```

You should see a table of age-specific fertility rates in 5 year age
groups and 1 year periods.

To check to see if this makes sense and is working right, let's
calculate the period Total Fertility Rate and see what's happened to
it in Italy and what is projected.

The definition of the TFR is the sum of the age specific rates. Since we have $n=5$ years, we multiply by $n$ to get the right number of person-years of exposure.

Here's the formula:
$$
TFR(t) = \sum_{x = 15}^{x = 45} nFx(t) * n
$$

And in "R", 
```{r}
TFR_t = colSums(nFxt)*5
```
Plotting fertility over time,
```{r}
plot(names(TFR_t), TFR_t, type = 'o', col = "blue", ylim = c(0, 3))
grid()
abline(v = 1995)
text(x = 1995, y = 2.5, "Replacement migration\n scenarios start\n in 1995", pos = 4)
```

We can see that Italy has a history of sub-replacement fertility, with
the TFR falling below 2 in about 1980. The UN forecasts predict the
TFR will rise slowly to 1.5 over the long-term.

# Mortality / Survival

For mortality and survival we'll focus on the the person-years lived
$nLx$ column of the lifetable, as this is all we'll need for
projection.

```{r}
## we repeat what we did for fertility nFx with mortality nLx
library(data.table)
##
dt <- fread("./nLx.csv") ## we'll use the name "dt" as a placeholder
## restrict to one country
mydt <- dt[Location == my_location]
head(mydt)
##
## and finally we'll create a matrix using the xtabs() functions with formula interface
nLxt <- mydt[, xtabs(nLx ~ x + t)] ## produces a table with rows age x and columns period t.
##
print(nLxt[1:5, 1:5])
```

(Note: the beginning age groups start out at 0-1, and 1-5. Our code
for making projection matrices will combine these groups to be 0-5.)

How could we calculate life expectancy easily from this matrix in
order to check the data? Hint: try the colSums() function.

# Age structure

Our files for age-structure include the entire history and the 2022
projections. (We read all of these in --  but for our analysis
we'll just use the starting population in the period we're interested
in and do the projection ourselves.)

```{r}
library(data.table)
dt <- fread("./nKx.csv") ## we'll use the name "dt" as a placeholder
## restrict to one country
mydt <- dt[Location == my_location]
head(mydt)
## and finally we'll create a matrix using the xtabs() functions with formula interface
nKxt <- mydt[, xtabs(nKx ~ x + t)] ## produces a table with rows age x and columns period t.
print(nKxt[1:5, 1:5])
```

Let's check the units by looking at total population size. Italy
should have something like 60 million people in 2020.

```{r}
Kt = colSums(nKxt)
million = 10^6
plot(names(Kt), Kt/million, col = "green")
title(paste(my_location, "population size"))
```

Looks like we have the right units, right?

# Analysis of starting age-structure

The UN's approach to replacement migraiton is to calculate the number
of immigrants that are needed in order to keep some aspect of the
population constant. They get the most dramatic results when the try
to keep the age-structure at working and older ages constant (as we'll
see in Part 2 of our exercise.

But before we get to replacement migration, let's start by look at
age-structure of the 1995 population that the UN report is trying to
keep constant.

Our population is  both sexes, but can still plot it as a kind of 'half-pyramid':

```{r}
nKx_zero = nKxt[,"1995"]/million
barplot(nKx_zero, horiz = T, xlab = "pop in millions", ylab = "age")
title(paste(my_location, "age-structure in 1995"))
```

Does this look like a stable population? Hard to say. It has shrinking
cohort sizes below about age 30, so it could be close to the stable
age-structure of a shrinking population. But the history of higher
fertility in the not-so-distant past might mean that there are
relatively few older people. Remember that fertility rates in the
1950s and 1960s were a lot higher than more recently.

The UN uses the "potential old-age support ratio" (OASR) as a measure
of age-structure. This is the ratio of potential workers per elderly
person, wich they define as the number aged 15-64 per person aged 65+.

Let's calculate the OASR for this population
```{r}
get_old_age_support_ratio <- function(nKx, x = seq(0,100, 5)) 
{
  workers = sum(nKx[x %in% 15:64])
  old = sum(nKx[x > 64])
  young = sum(nKx[x < 15])
  oasr = workers/old
  return(oasr)
}

oasr_zero = get_old_age_support_ratio(nKx_zero) # we're using "zero" to denote our starting population
print(oasr_zero)
```
  
So there are about 4 people of working age per person over age 65.

How does this compare to a stable population? The answer depends on
how we phrase the question. One way to answer this is to see what
constant fertility rate is required to have the same old-age support ratio as
observed, given some constant survival schedule. Let's try this using the
period survival schedule of 1995.

```{r}
## get the nLx's for 1995, with 5 year age groups for all groups.
nLx_raw = nLxt[,"1995"] 
nLx = c(sum(nLx_raw[1:2]), nLx_raw[-(1:2)]) ## this combines the 0- and 1-4 age groups.
```

Our stable age-structure formula (as you will study in days 1 and 2 of the workshop) is
$$
nCx = b * nLx * exp(-x * r),
$$
where 
* $b$ is the stable birth rate
* $nLx$ is the person-years lived aged x to x+n
* $r$ is the intrinsic rate of natural increase.

Let's find the growth "r" which gives us the an old-age support ratio
of about 4.1. We'll do this by "search", calculating OASR for a
range of values of "r".

```{r}
r.vec = seq(-.02, .03, .001)
oasr.vec <- NULL
x = seq(0, 100, 5)
for (i in 1:length(r.vec))
{
  this_r = r.vec[i]
  this_oasr = get_old_age_support_ratio(nLx * exp(-x*this_r))
  ## note: we can leave out b for the ratio because it cancels in the numerator and denominator
  oasr.vec[i] = this_oasr ## save result
}  
print(cbind(r.vec, oasr.vec))
```

So it looks like a stable growth rate $r$ of about 0.7 percent would
produce the same old-age support ratio over the long-term as we
observe in 1995.

The usual approximation of the growth rate from the TFR is

$$
r \approx \log (TFR * proportionFemale) / generationLength
$$

Inverting this, gives

$$
TFR \approx {\exp(r*generationLength) \over proportionFemale}
$$

```{r}
my_r = 0.7/100
my_generation = 30
my_prop_female = 0.4886
my_approx_TFR = exp(my_r * my_generation)/ my_prop_female
print(my_approx_TFR)
```

So it looks like we would need a TFR quite close to what was actually
observed in the 1950s or so.

# First set of exercises

Repeat our above analysis for "South Korea" and answer the following
questions.

(Hint: Do this by changing the variable "my_location" to "Republic of
Korea")

1. How would you describe the history of fertility in South Korea?
a) just like in Italy
b) higher fertility in past and lower fertility today, compared to Italy
c) lower fertility always than Italy

3. What is Korea's old-age support ratio in 1995. How does it compare
to Italy's at the same time?

4. What stable growth rate (and TFR) would be needed to keep that the
old-age support ratio constant at the level of South Korea in 1995
(with 1995 survival rates)?

5. How does this level of TFR (needed to maintain the 1995 old-age
support ratio) compare to South Korea's historical experience?

6. Why would demographers say that South Korea in 1995 is in the midst
of its "demographic dividend"? (Hint: look at the age-pyramid).

> Congratulations! You've reached the end of our first notebook, and
>  the end of our first set of exercises.








