---
title: "Reproductive values"
author: "Monica Alexander and Robert Chung"
output: html_document
---

```{r setup, include=FALSE}
# run this chunk first!
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

In this lab we're looking at how much each age group contributes to the eventual population size. First, read in the data and define our Leslie matrix and population projection functions:

```{r}
d <- read.csv("./projection_data_long.csv")
# alter the nLx and nFx columns so we dont have to worry about units later
d <- d %>% 
  mutate(nLx = nLx/10^5, nFx = nFx/1000)

```

```{r}
# define ffab 
ffab <- 0.4886
# define age groups
ages <- seq(0, 45, by = 5)
ages
```

```{r}
## function to define Leslie matrix, based on nLx and nFx values
leslie <- function(nLx,
                   nFx, 
                   n_age_groups=10,
                   ffab = 0.4886){
  L = matrix(0, nrow = n_age_groups, ncol = n_age_groups)
  L[1,] = ffab * nLx[1]*(nFx[1:n_age_groups]+nFx[2:(n_age_groups+1)]*nLx[2:(n_age_groups+1)]/nLx[1:n_age_groups])/2 # top row 
  diag(L[2:n_age_groups,1:(n_age_groups-1)]) = nLx[2:n_age_groups] / nLx[1:(n_age_groups-1)] # subdiagonal
  return(L)
}
```

```{r}
# Population projection function
# make a projection function, based on code in Lab 1
pop_project <- function(A, # Leslie matrix
                        initial_pop, 
                        n_age_groups = 10,
                        n_projections = 50){
  
  K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
  K[,1] <- initial_pop[1:n_age_groups]
  
  # do the projection!
  for(i in 2:(n_projections+1)){
    K[,i] <- A%*%K[,i-1] 
  }
  
  # get K in a form that's easier to ggplot
  # make into a dataframe, name columns as years and add an age column
  Kdf <- as.data.frame(K)
  colnames(Kdf) <- seq(from = 2015, to = (2015+n_projections*5), by = 5)
  Kdf <- cbind(age = seq(from = 0, to = 45, by = 5), Kdf)
  
  # get in long format and then add proportion of population in each age group
  Klong <- Kdf %>% 
    gather(year, population, -age) %>%
    group_by(year) %>%
    mutate(proportion = population/sum(population), 
           age = as.factor(age)) %>%
    ungroup()
  
  return(Klong)
}

```

## Projection for Senegal

In this lab, we're going to be using Senegal as an example. First, let's do a projection as usual and look at the total population. 

```{r}
nLx <- d$nLx[d$country=="Senegal"]
nFx <- d$nFx[d$country=="Senegal"]

A <- leslie(nLx = nLx, 
              nFx = nFx)

initial_pop <- d$nKx[d$country=="Senegal"]

project_results <- pop_project(A, initial_pop)
```

Plot the total population: we can see that Senegal has a positive growth rate. 

```{r}
tot_pop <- project_results %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>%
  mutate(R = c(NA, log(pop[2:n()]/pop[1:(n()-1)])/5))

ggplot(data = tot_pop, aes(x = year, y = pop, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Senegal - total population over time")
```

Make a record of the final population size:

```{r}
final_pop <- project_results$population[nrow(project_results)]
final_pop
```

## If we add people to certain age groups, what happens to the eventual population?

What if the first age group had 100 more people? Do you expect the final population size at the end of the period to be smaller, larger or the same size?

```{r}
initial_pop
```


```{r}
initial_pop_add_x <- initial_pop
initial_pop_add_x[1] <- initial_pop_add_x[1]+100
initial_pop_add_x
```

```{r}
project_results_add_x <- pop_project(A, initial_pop_add_x)
final_pop_add_x <- project_results_add_x$population[nrow(project_results_add_x)]
final_pop_add_x
```


Now try adding 100 people to each age group separately, and plot the results. Which age group results in the largest eventual population when 100 additional people are included?

```{r}
length(ages)
```


```{r}
## skeleton code
## here's a dataframe that you can fill in 
final_pop_sizes <- data.frame(age = ages, population = NA)

# you might want to do a for loop to add 100 people to each age group sequentially. 
for(i in 1:length(ages)){
  initial_pop_add_x <- initial_pop
  initial_pop_add_x[i] <- initial_pop_add_x[i]+100
  project_results_add_x <- pop_project(A, initial_pop_add_x)
  final_pop_add_x <- project_results_add_x$population[nrow(project_results_add_x)]
  final_pop_sizes$population[final_pop_sizes$age==ages[i]] <- final_pop_add_x
 }

final_pop_sizes
# plot final_pop_sizes by age 

```

```{r}
ggplot(data = final_pop_sizes, aes(x=age, y=population))+
  geom_line()+
  ggtitle("Population size after 50 projections adding 100 people to each age group - Senegal - BFDW 2017")
```



## Reproductive Values

We can get the reproductive values of the population directly from the Leslie matrix. Remember that the eigendecomposition of the Leslie matrix $A$ is 
$$
A = V \Lambda V^{-1}
$$
We got the stable age structure from the first column of the right-hand vector. We can get the reproductive values from the first row of the right-hand vector. To do this, we need to take the inverse of the eigenvector matrix. We also normalize to set the reproductive value of the first age group to equal 1. 

```{r}
# make function to get reproductive value

reproductive_value <- function(A # Leslie matrix
                               ){
  V <- eigen(A)$vector
  Vinv <- solve(V)
  Vstar <- diag(1/Vinv[,1]) %*% Vinv # normalize
  return(Re(Vstar[1,]))
}
```

Get the reproductive values for Senegal and plot over age. 

```{r}
rv.A <- reproductive_value(A)

## plot reproductive values
plot.df <- data.frame(age = ages, rv = rv.A)
ggplot(plot.df, aes(x = age, y = rv, group = 1))+
  geom_line() + geom_point() + ggtitle("Reproductive values for Senegal")
```

## What happens if mortality goes up?

```{r}
nLx <- d$nLx[d$country=="Senegal"]
nFx <- d$nFx[d$country=="Senegal"]

nLx.lower <- nLx*0.5

A.lower <- leslie(nLx = nLx.lower, 
              nFx = nFx)

rv.A.lower <- reproductive_value(A.lower)

## plot reproductive values
plot.df <- data.frame(age = ages, rv = rv.A.lower)
ggplot(plot.df, aes(x = age, y = rv, group = 1))+
  geom_line() + geom_point() + ggtitle("Reproductive values for Senegal")

```



Can do this for all countries in the dataset. 

```{r}
# loop through countries, get stable age and rv
ages <- seq(0, 45, by = 5) # define age groups

stable_df <- c() # initialize an empty object to store results

for(country in unique(d$country)){
  #leslie matrix for current country
  A <- leslie(nLx = d$nLx[d$country==country], nFx = d$nFx[d$country==country])
  rv.A <- reproductive_value(A)
  # put results in a dataframe
  df <- data.frame(age = ages, country = country, rv = rv.A)
  # combine this dataframe with overall df
  stable_df <- rbind(stable_df, df)
}

head(stable_df)
```
```{r}
# plot reproductive values by country
ggplot(data = stable_df, aes(x = age, y = rv, color = country)) + 
  geom_line()+
  ylab("reproductive value")+
  ggtitle("Reproductive values by country")

```