---
title: "Lab 1. Population projection: data exploration"
author: "Monica Alexander and Robert Chung"
output: html_document
---

```{r setup, include=FALSE}
# run this chunk first!
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
tot = 0
```

We will be exploring some of the concepts on population projection from the session this morning. The aim of this lab is to introduce you to the dataset that we'll be using.

The dataset has values for $_nL_x$ (person-years lived between ages $x$ and $x+n$), $_nF_x$ (age-specific fertility) and $_nK_x$ (population between ages $x$ and $x+n$) for six countries, which have different types of mortality, fertility and age structures:

- Australia 
- Kenya 
- Singapore 
- Niger 
- Russia
- Senegal

Data are for females in 2015 and were obtained from the [World Health Organisation](http://apps.who.int/gho/data/node.main.687?lang=en) ($_nL_x$) and [United Nations Population Division](https://esa.un.org/unpd/wpp/) ($_nF_x$ and $_nK_x$). 

This exercise uses tidyverse coding conventions, so we loaded the tidyverse package. If you haven't already done that, you should do that now. You should also have the data file "projection_data_long.csv" in your working directory.

```{r}
d <- read.csv("projection_data_long.csv")
```

Throughout this lab, we'll be making use of the `%>%` operator. This is like a 'pipe', which takes whatever is on the left hand side and feeds it forward into a function. For example, below we are taking our data `d`, filtering it to only include Australia, and then printing the head (the first six rows).

```{r}
d %>% 
  filter(country=="Australia") %>% 
  head
```

Notice that:

- Although we usually use a radix of 1.0, in these data the magnitude of $_nL_x$ values suggests a radix of 100,000 (that is, 10^5)
- $_nF_x$ is the age-specific birth rate per 1,000 women
- $_nK_x$ is age-specific population in 000s

### Person-years lived: $_nL_x$

Let's investigate the values of $_nL_x$ by country. First, let's calculate the life expectancy at birth, $e_0$, for all countries. Remember that $$e_0 = \frac{\sum {_nL_x}}{l_0}$$

To calculate in R we can use `group_by` to do the calculation separately for each country.

```{r}
# calculate life expectancy at birth 
# we know radix is 100,000
radix <- 10^5

d %>% 
  group_by(country) %>% 
  summarise(e0 = sum(nLx)/radix) %>%
  arrange(e0) # arrange by increasing life expectancy

```

There is more than 23 years difference in the life expectancy at birth for females in Niger compared to Singapore. 

### Fertility rates: $_nF_x$
We can use the age-specific fertility rates to calculate some summary fertility measures for each country. Note that we will assume that the proportion of babies that are female (fraction of females at birth) is 0.4886. If we had additional information, we could calculate the exact Ffab, but if not we'll use the default value. Using the information we have, we can calculate the total fertility rate (TFR), gross reproduction ratio (GRR) and net reporduction ratio (NRR). Remember that
$$ TFR = \sum {_nF_x} \cdot n$$
$$ GRR = TFR \cdot f_{fab}$$
$$NRR = \frac{\sum {_nF_x}\cdot {_nL_x} \cdot f_{fab}}{l_0} $$
For these data, we will limit the age range for our calculations to 15 to 49. 

```{r}
# calculate TFR, GRR and NRR by country
# set fraction female at birth to be 0.4886
ffab <- 0.4886

d %>% 
  filter(age >= 15, age < 50) %>% # restrict our dataset to be fertility rates for women aged 15-49
  group_by(country) %>% 
  summarise(tfr = sum(nFx*5)/1000, 
            grr = tfr*ffab,
            nrr = sum(nFx*nLx*ffab/radix)/1000) %>%
  arrange(tfr)
```

We can also plot the age-specific fertility rates for each country. 

```{r}
# create a dataset which only has reproductive ages (15-49)
d_rpa <- d %>% filter(age>=15, age <50)

# plot age-specific fertility rates
ggplot(data = d_rpa, aes(x = age, y = nFx, color = country)) + 
  geom_line()+
  ylab("Age-specific fertility rate (births/1000 women)")+
  ggtitle("Age-specific fertility rate by country")+ 
  scale_color_brewer(palette="Set1")
```


### Population age structure: $_nK_x$
Finally, let's look at the age structure of each country. We can calculate the mean age:

```{r}
# calculate mean age 
d %>% 
  group_by(country) %>%
  summarise(mean_age = sum((age+2.5)*nKx)/sum(nKx)) %>%
  arrange(mean_age)
```

We can also plot the proportion of the population in each age group. This is like half of a population pyramid (the female half). First we need to calculate the proportion in each age group from the raw numbers.

```{r}
# let's calculate the proportion in each age group so we can compare across countries
d <- d %>% 
  group_by(country) %>%
  mutate(nCx = nKx/sum(nKx))
```

```{r}
ggplot(d, aes(x = age, y = nCx)) + 
  facet_wrap(~country, ncol=3)+
  geom_bar(stat="identity", position = "dodge")+
  ggtitle("Proportion of population in each age group")+
  ylab("proportion")+
  coord_flip()
```


## Questions to think about and check your understanding


**Q1. $_5L_{10}$ means:**

A. The number of person years lived between ages 5 and 10

B. The number of person years lived between ages 10 and 15

C. The population between ages 5 and 10


**Q2. Evaluate this statement: "$_5L_{5}$ can never be larger than $_5L_{0}$".**

A. True

B. False

C. Depends on level of infant mortality


**Q3. The Gross rate of reproduction (GRR) is** 

A. Always less than or equal to the Net Rate of Reproduction (NRR)

B. Always more than or equal to the NRR

C. Similar magnitude to NRR, but relationship depends on mortality 

