---
title: "Lab: Population projection and stable populations"
author: "Monica Alexander and Robert Chung"
output: html_document
---

```{r setup, include=FALSE}
# run this chunk first!
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
tot = 0

```

In this lab we will do a population projection exercise using the fertility, mortality and population data from the dataset explored in Lab 1: Exploration. We will then compare these results to the growth rate and population structure implied by stable population theory. 

## Read in data

We are using the same dataset from Lab: Exploration, which had fertility, mortality and population data for six different countries. Before using the data, change the units of $_nL_x$ to have a radix of 1 and $_nF_x$ to be births per woman. As you may recall, the original $_nL_x$ had a radix of 100000 (10^5) and the original $_nF_x$ were per thousand, so we'll divide them by 10^5 and 1000, respectively.

```{r}
d <- read.csv("./projection_data_long.csv")
# alter the nLx and nFx columns so we dont have to worry about units later
d <- d %>% 
  mutate(nLx = nLx/10^5, nFx = nFx/1000)
```

## Leslie matrices

We need to create a Leslie matrix which contains the fertility and survivorship information in order to project a population forward. Remember that elements in the top line of the Leslie matrix are equal to 
$$
_nL_0 \cdot \frac{1}{2}(_nF_x + {_nF_{x+n}}\cdot\frac{_nL_{x+n}}{_nL_x}) \cdot F_{fab}
$$
and the subdiagonals are equal to
$$
\frac{_nL_{x+n}}{_nL_x}
$$
Below is a function that helps to create a Leslie matrix. It takes four inputs: 

- a vector of $_nL_x$ values
- a vector of $_nF_x$ values
- the number of age groups. By default this is 10, because we are considering the five year age groups from ages 0--50.
- The fraction female at birth, which is set to a default value of 0.4886. 

With these inputs, the function `leslie` creates a square matrix with dimension equal to `n_age_groups` and fills in the top row and subdiagonal according to the equations above. 

```{r}
## function to define Leslie matrix, based on nLx and nFx values
leslie <- function(nLx,
                   nFx, 
                   n_age_groups=10,
                   ffab = 0.4886){
  L = matrix(0, nrow = n_age_groups, ncol = n_age_groups)
  L[1,] = ffab * nLx[1]*(nFx[1:n_age_groups]+nFx[2:(n_age_groups+1)]*nLx[2:(n_age_groups+1)]/nLx[1:n_age_groups])/2 # top row 
  diag(L[2:n_age_groups,1:(n_age_groups-1)]) = nLx[2:n_age_groups] / nLx[1:(n_age_groups-1)] # subdiagonal
  return(L)
}
```

We can use this function to create a Leslie matrix using the $_nL_x$ and $_nF_x$ values for Australia:

```{r}
LAA <- leslie(nLx = d$nLx[d$country=="Australia"], 
              nFx = d$nFx[d$country=="Australia"])
LAA
```

We could also use $_nL_x$ and $_nF_x$ values for any combination of the different countries, for example, $_nL_x$ values from Australia and $_nF_x$ from Senegal:

```{r}
LAS <- leslie(nLx = d$nLx[d$country=="Australia"], 
              nFx = d$nFx[d$country=="Senegal"])
```


## Population projection
We can now use the Leslie matrices to project a population forward in time. Let's start with the Australia-only Leslie matrix and project forward Australia's 2015 population 250 years. Note that because we are using five-year age groups, the population projection happens in five-year steps. 

```{r}
n_age_groups <- 10 # 0-50 in 5yr age groups
n_projections <- 50 # want to project forward 50*5 = 250 years
initial_pop <- d$nKx[d$country=="Australia"] # set initial population to be Australia's population in 2015
# define population matrix K
K <- matrix(0, nrow = n_age_groups, ncol = n_projections+1)
K[,1] <- initial_pop[1:n_age_groups]

# do the projection!
for(i in 2:(n_projections+1)){
  K[,i] <- LAA%*%K[,i-1] 
}

head(K[,1:5])
```

Now we have a matrix of populations by age for Australia projected into the future. We can use these numbers to investigate:

- The total population over time
- The share of the population in each age group over time
- The population growth rate over time

First, let's get the matrix `K` in a form that's a bit easier to plot. This involves making `K` into a dataframe in 'long' format, so that every row refers to a different year and age combination. 

```{r}
# get K in a form that's easier to ggplot
# make into a dataframem name columns as years and add an age column
Kdf <- as.data.frame(K)
colnames(Kdf) <- seq(from = 2015, to = (2015+n_projections*5), by = 5)
Kdf <- cbind(age = seq(from = 0, to = 45, by = 5), Kdf)

# get in long format and then add proportion of population in each age group
Klong <- Kdf %>% 
  gather(year, population, -age) %>%
  group_by(year) %>%
  mutate(proportion = population/sum(population), 
         age = as.factor(age))
head(Klong)
```
Create a data frame that has the total population by year, and calculate the annual growth rate based on the equation

$$
K(t+5) = K(t) \cdot e^{5R}
$$
So
$$
R = \frac{1}{5} \cdot \log \frac{K(t+5)}{K(t)}
$$

```{r}
# total population by year
tot_pop <- Klong %>% 
  group_by(year) %>%
  summarise(pop = sum(population)) %>%
  mutate(R = c(NA, log(pop[2:n()]/pop[1:(n()-1)])/5))
head(tot_pop)
```

Now we can plot total population, growth rate, and proportion in each age group over time.  

```{r}

# plot total population over time 
ggplot(data = tot_pop, aes(x = year, y = pop, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Total population over time")

# plot growth rate over time
ggplot(data = tot_pop, aes(x = year, y = R, group = 1)) +
  geom_point() + geom_line()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Growth rate of population over time")

# plot proportion in each age group over time
ggplot(data = Klong, aes(x = year, y = proportion, group = age, color = age)) +
  geom_point() + geom_line()+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+
  ggtitle("Proportion of population in each age group over time")
```


## Stable population quantities 

We can do an eigen-decomposition of the Leslie Matrix to get the values of the stable growth rate $r$ and stable population age structure. If we compare these values to the values of the growth rate $R$ and population age structure in the final projection period, you can see that they are very similar. 

```{r}
eigen.AA <- eigen(LAA)
r.AA <- log(Re(eigen.AA$value[1]))/5 
v <- Re(eigen.AA$vectors[,1])
k.AA <- v/sum(v)

# compare stable population rate and rate at end of projection period
cbind(stable = r.AA,  proj = tot_pop$R[nrow(tot_pop)])

# compare stable population age distribution and age distribution at end of projection period
cbind(stable = k.AA, proj = Klong$proportion[Klong$year==(2015+n_projections*5)])

```

```{r}
ages <- seq(0, 45, by = 5)
ggplot(data = data.frame(age = ages, proportion = k.AA), aes(x = age, y = proportion))+
  geom_bar(stat = 'identity')+
  coord_flip()+
  ggtitle("Stable age population structure")
```


## Exercises

1. Would you expect a population projection for Niger to converge to the stable properties more quickly or slowly than Australia? Why? Repeat the projection exercise for Niger to see what happens.
2. Investigate how different fertility and mortality situations effect the stable population characteristics. Obtain eigendecompositions of different combinations of high and low fertility/mortality countries and look at the implied growth rates and population age structures. 


## Questions to think about and check your understanding

**Q1. What does Australia's stable growth rate tell you about Australia's NRR?**

**Q2. True or False: If everything else remained the same, doubling the age-specific fertility rates at every age would double the Total Fertility Rate.**

**Q3. True or False: If everything else remained the same, doubling the age-specific fertility rates at every age would double the growth rate r.**

**Q4. The growth rate r obtained from the eigendecomposition of the Leslie matrix can be interpreted as:**

A. The current growth rate of the population

B. The current growth rate of the population, ignoring migration

C. The stable growth rate of a population; i.e. the growth rate it will converge to if fertility and mortality rates remained constant over time.


